---
title: "API 101"
author: "Edzer Pebesma"
date: "Aug 23, 2023"
---

# Why API 101?

There is a lot of talking about APIs. What is it?

* API stands for Application Programming Interface. This is completely generic.
* Any software consists of functions (or methods) that do something. What they do is explained
  in the documentation. If the documentation and the implementation are not in correspondence,
  then we call it a bug (in either the function or the documentation, or both)
* Defining an API means: 
    * defining how the function is executed:
        * choosing a name for the function, 
        * choosing names for input arguments, and defining what they should be (their type)
        * the calling protocol (e.g. GET or POST)
    * what the function shall do (usually: what it returns)
    * writing this down
* Implementing an API means writing working code that adheres to the defined API

An API may be as simple as an R function:

```{r}
add_three = function(a, b, c) {
	stopifnot(is.numeric(a), is.numeric(b), is.numeric(c))
	a + b + c
}
add_three(4, 5, 6)
```
R is a weakly typed language, so the function definition does not
prescribe which type `a`, `b` and `c` shall have, therefore we can /
need to check their type inside the function.

# Web APIs

Web APIs define how the web works: when we click something in a web page a URL is sent to a web server, and a response is given; the browser reacts to that response.

Web APIs can be easily create by either R or Python.

## By R

Using package [plumber](https://www.rplumber.io/), we can create an API definition as follows:

```
# plumber.R

#* Echo back the input
#* @param msg The message to echo
#* @get /echo
function(msg="") {
  list(msg = paste0("The message is: '", msg, "'"))
}

#* Plot a histogram
#* @serializer png
#* @get /plot
function() {
  rand <- rnorm(100)
  hist(rand)
}

#* Return the sum of two numbers
#* @param a The first number to add
#* @param b The second number to add
#* @post /sum
function(a, b) {
  as.numeric(a) + as.numeric(b)
}
```
and run this by
```{r eval=FALSE}
pr("plumber.R") |> pr_run(port=8000)
```
When doing this, we see
```
Running plumber API at http://127.0.0.1:8000
Running swagger Docs at http://127.0.0.1:8000/__docs__/
```
indicating 

* where the service is running
* where their documentation is found

[ do it; show; try GET on the POST, curl --data POST ]

## By Python

A possibility to create a web service with Python is using Flask; 
save the following

```
from flask import Flask

app = Flask(__name__)

@app.route("/")
def hello_world():
    return "<p>Hello, World!</p>"
```
to `hello.py`, and run
```
export FLASK_APP=hello.py
flask run
```
this then prints:
```
 * Serving Flask app 'hello.py' (lazy loading)
 * Environment: production
   WARNING: This is a development server. Do not use it in a production deployment.
   Use a production WSGI server instead.
 * Debug mode: off
 * Running on http://127.0.0.1:5000/ (Press CTRL+C to quit)
```
and on that web address we can interact with the web service:

* `curl  http://127.0.0.1:5000/` prints `<p>Hello, World!</p>`
* opening the address `http://127.0.0.1:5000/` in a web browser shows the html 

