---
title: "Using an openEO back-end from R"
author: "Edzer Pebesma"
date: "Aug 23, 2023"
---

# openEO is an API

Well, yes, and along with that 

* a software ecosystem with client and server implementations that makes it work, 
* a number of large-scale, cloud-based deployments, both public and private
* a community of people working on it and using it

## Where is it?

From https://openeo.org/, software, Specifications:

* [openEO API](https://github.com/Open-EO/openeo-api): source code of the generic aspects: end points
* [openEO processes](https://github.com/Open-EO/openeo-processes): source code of all the processes that form a process graph
* [link to the rendered Swagger docs](https://api.openeo.org/)

How does it work?

1. Connecting to a server gives you the ability to query:
    * image collections available
    * processes supported
    * file formats supported for output
2. Given that info, one can iteratively build instructions: with image collection `x`, select region `y` and period `z`, and do this, that and then that with it.
   This set of instructions is arbitrary complex.
3. After `login()`, you can send a the set of instructions ("process graph") to the server, and either
    * wait for an answer (synchronous)
    * go for coffee, or lunch, or check impatiently whether the job has finished (asynchronous)


## Examples

### Example 1: using OpenEO Platform

From https://r.geocompx.org/gis.html?q=openeo#openeo

```
plots_json <- geojson_read("plots.geojson")
```

```{r echo=FALSE, message=FALSE}
if (file.exists("con.RData"))
	load("con.RData")
library(openeo)
active_connection(con)
```

```{r eval=!exists("con")}
library(openeo)
con = connect("https://openeo.cloud")
# Login, see https://docs.openeo.cloud/getting-started/r/#authentication
login()
```

```{r echo=FALSE}
save(list = "con", file = "con.RData")
```

User support platform: https://discuss.eodc.eu/c/openeo-platform/5

```{r}
p = processes() # load available processes
collections = list_collections() # load available collections
formats = list_file_formats() # load available output formats
# Load Sentinel-2 collection
s2 = p$load_collection(
  id = "SENTINEL2_L2A",
  spatial_extent = list(
    west = 7.9, east = 8.1,
    north = 51.1, south = 50.9
  ),
  temporal_extent = list("2021-01-01", "2021-01-31"),
  bands = list("B04", "B08")
)
# Compute NDVI vegetation index
compute_ndvi = p$reduce_dimension(
  data = s2, dimension = "bands",
  reducer = function(data, context) {
    (data[2] - data[1]) / (data[2] + data[1])
  }
)
# Compute maximum over time
reduce_max = p$reduce_dimension(
  data = compute_ndvi, dimension = "t",
  reducer = function(x, y) {
    max(x)
  }
)
# Export as GeoTIFF
result = p$save_result(reduce_max, formats$output$GTiff)
# Execute processes
f = compute_result(graph = result, output_file = tempfile(fileext = ".tif"))
library(stars)
read_stars(f) |> plot()
```

# A larger example, using Copernicus Data Space Ecosystem

From: https://r.iresmi.net/posts/2023/copernicus_openeo_ndvi_time_series/

```{r echo=FALSE, message=FALSE}
if (file.exists("cnx.RData"))
	load("cnx.RData")
library(openeo)
active_connection(cnx)
```

```{r}
library(openeo)
library(signal)
library(tidyverse)
library(sf)
library(geojsonio)
library(janitor)
library(fs)
```

```{r}
# 3 sample plots with field `landcover`: "agriculture", "forest", "city"
if (file_exists("plots.geojson")) file_delete("plots.geojson")
plots <- read_sf("plots.gpkg", layer = "plots") |>
  write_sf("plots.geojson")
plots_json <- geojson_read("plots.geojson")

# temporal span (start - end)
date_span <- c("2020-01-01", "2022-12-31")

# Area name (for the final graphic plot)
location <- "Bugey"
```

```{r eval=!exists("cnx")}
cnx <- connect(host = "https://openeo.dataspace.copernicus.eu")

# Your console will require you to paste a code and your browser will ask to
# grant access
login()
```
```{r echo=FALSE}
save(list = "cnx", file = "cnx.RData")
```

```{r eval=interactive()}
list_collections()
describe_collection("SENTINEL2_L2A")

# List of available openEO processes with full metadata
processes <- list_processes()

# List of available openEO processes by identifiers (string)
print(names(processes))
process_viewer(processes)

# print metadata of the process with ID "load_collection"
print(processes$load_collection)

# get the formats
formats <- list_file_formats()
```

```{r}
# get the collection list to get easier access to the collection ids,
# via auto completion
collections <- list_collections()

# get the process collection to use the predefined processes of the back-end
p <- processes()

# compute NDVI
ndvi <- function(data, context) {
  red <- data[1]
  nir <- data[2]
  (nir - red) / (nir + red)
}

# remove pixels not classified as vegetation (4) or non-vegetation (5), i.e.:
# water, shadows, clouds, unclassified...
filter_unusable <- function(data, context) {
  scl <- data[3]
  !(scl == 4 | scl == 5)
}

# check processing status
status_job <- function(job) {
  while (TRUE) {
    if (!exists("started_at")) {
      started_at <- ymd_hms(job$created, tz = "UTC")
      message(capabilities()$title, "\n",
              "Job « ", job$description, " », ",
              "started on ",
              format(started_at, tz = Sys.timezone(), usetz = TRUE), "\n")
    }

    current_status <- status(job)
    if (current_status == "finished") {
      message(current_status)
      break
    }

    current_duration <- seconds_to_period(difftime(Sys.time(),
                                                   started_at,
                                                   units = "secs"))
    message(sprintf("%02d:%02d:%02d",
                    current_duration@hour,
                    minute(current_duration),
                    floor(second(current_duration))), " ",
            current_status, "...")
    Sys.sleep(30)
  }
}
```

```{r}
s2_dataset <- p$load_collection(id = collections$SENTINEL2_L2A,
                            temporal_extent = date_span,
                            bands = c("B04", "B08", "SCL"))

mask <- p$reduce_dimension(data = s2_dataset,
                           dimension = "bands",
                           reducer = filter_unusable)

result <- s2_dataset |>
  p$mask(mask) |>
  p$reduce_dimension(dimension = "bands",
                     reducer = ndvi) |>
  p$aggregate_spatial(geometries = plots_json,
                      reducer = mean) |>
  p$save_result(format = "CSV")
```

```{r eval=interactive()}
job <- create_job(graph = result,
                  title = "NDVI",
                  description = "NDVI plots")

start_job(job)
# list_jobs()
status_job(job)
# logs(job)

f <- download_results(job = job, folder = "results")
ndvi_data <- read_csv(f[[1]]) |> clean_names()
```

```{r eval=!interactive()}
ndvi_data <- read_csv("results/timeseries.csv") |> clean_names()
```

```{r}
# filtering (optional)
ndvi_filtered <- ndvi_data |>
  arrange(feature_index, date) |>
  drop_na(avg_band_0) |>
  group_by(feature_index) |>
  mutate(ndvi = sgolayfilt(avg_band_0, n = 5)) |>
  ungroup()

# plot
landcover_colors <- list("agriculture" = "goldenrod",
                         "forest" = "darkgreen",
                         "city" = "darkgrey" )

ndvi_filtered |>
  full_join(plots |>
              st_drop_geometry() |>
              mutate(feature_index = row_number() - 1)) |>
  ggplot(aes(date, ndvi,
             group = landcover,
             color = landcover,
             fill = landcover)) +
  geom_point(alpha = 0.2) +
  geom_smooth(span = 0.05) +
  scale_x_datetime(date_breaks = "year",
                   date_minor_breaks = "month",
                   date_labels = "%Y") +
  scale_y_continuous(breaks = scales::breaks_pretty()) +
  scale_color_manual(values = landcover_colors) +
  scale_fill_manual(values = landcover_colors) +
  guides(color = guide_legend(reverse = TRUE),
         fill = guide_legend(reverse = TRUE)) +
  labs(title = "Change of vegetation",
       subtitle = location,
       y = "NDVI",
       color = "Landcover",
       fill = "Landcover")

ggsave("results/ndvi.png")
```

## Third example, using Brian's openEO back end, self-hosted at AWS

The following example is taken from Brian Pondi,
https://github.com/PondiB/openeocubes

```{r}
library(openeo)

# connect  to the back-end when deployed locally
# con = connect("http://localhost:8000")
# connect  to the back-end when deployed on aws
con = connect("http://35.89.54.201:8000")

# basic login with default params
login(user = "user", password = "password")

# get the collection list
collections = list_collections()

# to check description of a collection
collections$`sentinel-s2-l2a-cogs`$description

# Check that required processes are available.
processes = list_processes()

# to check specific process e.g. filter_bands
if (interactive()) {
	describe_process(processes$filter_bands)
}

# get the process collection to use the predefined processes of the back-end
p = processes()

# load the initial data collection and limit the amount of data loaded
datacube_init = p$load_collection(id = "sentinel-s2-l2a-cogs",
                                  spatial_extent = list(west=416812.2,
                                                        south=5803577.5,
                                                        east=422094.8,
                                                        north=5807036.1),
                                  temporal_extent = c("2016-01-01", "2020-12-31"),
                                  # extra optional args for datacubes regularization -> courtesy of gdalcubes
                                  pixels_size = 10,
                                  time_aggregation = "P1M",
                                  crs = 32633)

# filter the data cube for the desired bands
datacube_filtered = p$filter_bands(data = datacube_init, bands = c("B04", "B08"))

# bfast custom change detection method (UDF - user-defined function, here in R):
change.detection = 'function(x) {
  knr <- exp(-((x["B08",]/10000)-(x["B04",]/10000))^2/(2))
  kndvi <- (1-knr) / (1+knr)
  if (all(is.na(kndvi))) {
    return(c(NA,NA))
  }
    kndvi_ts = ts(kndvi, start = c(2016, 1), frequency = 12)
    library(bfast)
    tryCatch({
        result = bfastmonitor(kndvi_ts, start = c(2020,1), level = 0.01)
        return(c(result$breakpoint, result$magnitude))
      }, error = function(x) {
        return(c(NA,NA))
      })
  }'

# run udf
datacube_udf = p$run_udf(data = datacube_filtered, udf = change.detection, names =  c("change_date", "change_magnitude"))

# supported formats
formats = list_file_formats()

# save as GeoTiff or NetCDF
result = p$save_result(data = datacube_udf, format = formats$output$NetCDF)

# Process and download data synchronously
start.time <- Sys.time()
compute_result(graph = result, output_file = "change_detection.nc")
end.time <- Sys.time()
time.taken <- end.time - start.time
time.taken
print("End of processes")
```

# show:
```{r}
library(stars)
#r = read_mdim("change_detection.nc", proxy = TRUE)
r = read_stars('NETCDF:"change_detection.nc":change_date')
class(r[[1]]) = "Date"
r["t"] = as.POSIXct(r[[1]])
class(r$t) = c("POSIXct", "POSIXt")
ra = range(r$t, na.rm = TRUE)
br = seq(ra[1], ra[2], length.out = 10)
plot(r["t"], breaks = br)

m = read_stars('NETCDF:"change_detection.nc":change_magnitude')
plot(m, breaks = "equal")
plot(m)
```
